Vagueness, Noise, and Signaling

### 2015-08-13 paper/version_04/paper.tex
first pre-final version 04

\begin{abstract}
  Signaling games are popular models for studying the evolution of meaning, but typical
  approaches do not incorporate vagueness as a feature of successful signaling.  Complementing
  recent like-minded models, we describe an aggregate population-level dynamic that describes a
  process of imitation of successful signaling behavior under imprecise perception and
  realization of similar stimuli. Applying this new dynamic to a generalization of Lewis'
  signaling games, we show that stochastic imprecision leads to vague, yet by-and-large
  efficient signal use, and, moreover, that it unifies evolutionary outcomes and helps avoid
  suboptimal categorization. \myred{The upshot of this is that generalization emerges, at an
    aggregate level, without agents actually generalizing.}
\end{abstract}

@@

A number of authors have recently tried to explain why vagueness evolved as something that is
itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing \citep[e.g.][]{FrankeJager2010:Vagueness-Signa} or the result of particular
low-level learning strategies \citep[e.g.][]{OConnor2013:The-Evolution-o}. This paper
contributes to the latter line of thought. Concretely, we introduce an dynamic that describes
average behavior resulting from imitation under confusability of similar stimuli. The dynamic
proposed here complements previous like-minded accounts. We show that stochastic noise can not
only lead to vague, yet relatively efficient signal use, but can also unify evolutionary
outcomes and help avoid suboptimal categorization. Moreover, the system shows generalization
behavior without a single generalizing agent.

@@

How does a tendency to confuse similar states interact with a drive towards more efficient
signaling strategies under the imprecise conditional imitation dynamic? To be able to properly
address this question, it pays to be more precise about similarity of states, the conditional
confusion probabilities $P_o$ and $P_r$ and the utilities of our sim-max games.

\subsection{Confusion \& utilities}
\label{sec:iterated-diffusion}

Confusion of states should intuitively be a function of their perceptual similarity: the more
similar two states are, the more likely they could be mistaken for each other.  To make this
concrete, let us assume that the state space of the sim-max game consists of $n \ge 2$ states
that are equally spaced across the unit interval, including $0$ and $1$. The distance
$\card{\mystate{i} - \mystate{j}}$ is the objective, physical similarity between two states
$\mystate{i}$ and $\mystate{j}$. Distance in physical space feeds into a perceptual similarity
function, as described by \citet{Nosofsky1986:Attention-Simil}:
\begin{align*}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \imprecision) =
      \begin{cases}
    1 & \textrm{if } \imprecision = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \imprecision = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \imprecision^2} \right ) & \textrm{otherwise,} \\
    \end{cases}
\end{align*}
where $\imprecision \ge 0$ is an imprecision parameter. When $\imprecision=0$ agents perfectly
discriminate between states; when $\imprecision \rightarrow \infty$ agents cannot discriminate
states at all. Figure~\ref{fig:NosofskySim} gives an impression of Nosofsky-similarity for
different parameter values. Other formalizations of perceptual similarity are possible,
including ones that allow for different discriminability in different areas of the state space,
but we stick with Nosofsky's similarity function for the time being, because it is
mathematically simple, yet an established notion in mathematical psychology.

@@

We further assume that the probability of confusing any two states $\mystate{i}$ and
$\mystate{j}$ is proportional to their perceived similarity and, to keep matters simple, that
observational confusability $P_o$ is simply the same as realizational confusability $P_r$ and
that both are governed by the same imprecision parameter $\alpha$:
\begin{align*}
  P_o(\mystate{o} \probbar \mystate{a}) & \propto \similarity(\mystate{o}, \mystate{a} \myts ; \myts
  \imprecision) &   P_r(\mystate{r} \probbar \mystate{i}) & \propto \similarity(\mystate{r}, \mystate{i} \myts ; \myts
  \imprecision) \,.
\end{align*}
For $\imprecision = 0$, we obtain trivial confusion probabilities, i.e., confusion probability
zero between non-identical states and our dynamic reduces to the replicator dynamics. For
$\imprecision > 0$, any state can be confused for any other state with some positive
probability. For the uninteresting case of $\imprecision \rightarrow \infty$, confusion is
maximal and every state can be perceived or realized as any other state with probability
$\nicefrac{1}{\card{\States}}$.

@@

In conclusion, to structure thinking about the behavior of the \icid, the system is governed by
two parameters: imprecision $\imprecision$ and tolerance $\toler$, which can vary
independently.

\subsection{Numerical simulations}
\label{sec:simulations}

The main question we should address is whether state confusability and utility-driven imitation
reasonably interact, and if so, whether the inclusion of confusability has any noteworthy
effects on the evolving meaning of signals. It transpires that moderate levels of imprecision
$\imprecision$ can lead to reasonably efficient, yet vague signaling behavior. Moreover,
imprecision can have further accelerating and, surprisingly, unifying effects on meaning
evolution. It appears that state confusability has an effect on aggregate signaling behavior
that can be described as generalization or systematization, as if ``conclusions'' about what
works for one state are ``carried over'' to similar states. This, however, is merely an
emergent phenomenon in the sense no single agent genuinely generalizes over stimuli or reasons
about what a more systematic signaling strategy would be.

For all simulations, priors were flat: $P_a(\state) = P_a(\state')$ for all
$\state, \state'$. We ran 50 trials of the discrete-time dynamic in (\ref{eq:nRD}), starting
with randomly sampled sender and receiver strategies, for each triplet of independent parameter
values: $\ns = \card{T} \in \set{6, 10, 50, 90}$,
$\imprecision \in \set{0, 0.05, 0.1, 0.2, 0.3}$, $\toler \in \set{0.05, 0.1, 0.2, 0.3}$. Each
trial ran for a maximum of 200 update steps of the \rdd. A trial was considered converged, and
thus stopped before the maximum of 200 rounds, if the total amount of change between strategies
before and after an \rdd step was smaller than a suitably chosen threshold. It is not
guaranteed that strategies at halting time had converged to the eventual attracting state,
whether they ran for 200 rounds or not. Our notion of convergence is therefore only a
categorical measure for reaching a certain (well-considered, but eventually arbitrary) degree
of stability. In principle, rather than setting a hard limit on the maximum number of rounds
and analyzing convergence categorically, one could also run each trial until reaching our
criterion for convergence and analyze the number of rounds it took to get there.  Our choice
was initially motivated by pragmatic concerns regarding length of simulation time, but it is
also justifiable if we reflect on the nature of language evolution as a continuous process.
Language adapts to an ever-changing reality; it is rare, if not impossible, that a lexicon has
an infinite amount of time to neatly converge to a stable state. Therefore, investigating speed
of convergence within a limited time frame can be seen as better reflecting the natural
constraints of language evolution.

@@

 Representative examples for resulting strategy pairs are given in
Figure~\ref{fig:example_strats}. Figure~\ref{fig:example_stratsA} shows a strategy pair at
stopping time with 90 states, tolerance $\toler = 0.1$ and imprecision $\imprecision = 0$. Zero
imprecision means that the trial was effectively an application of the plain \rd. Noteworthily,
the given sender strategy approximates a pure sender strategy that crisply partitions the state
space into non-convex sets. The irregular shape of the receiver strategy suggests that the
pictured strategy pair has not yet reached a dynamically stable state under the \rd. Indeed,
the trial was stopped for reaching the maximum of 200 rounds. In contrast, the outcome of a
trial with identical parameters, except with imprecision $\imprecision = 0.0.5$, which is shown
in Figure~\ref{fig:example_stratsB}, had converged (in our technical sense) after 99 rounds of
iterated \rdd. The sender strategy shows a smooth blending from one ``category'' to the other,
and the receiver's interpretations are rather extended curves, peaking at a central point in
the relevant ``categories.''

@@

\subsubsection{Dependent measures}

@@

\paragraph{Convexity.} At least for sender strategies, which develop
faster than receiver strategies, it also makes sense to define a
categorical measure of convexity that compensates for potential
vagueness. To determine whether a sender strategy $\Sstrat$ is convex
despite possibly being vague, we look at the derived pure strategy
$\Spure$ for which $\Spure(\state) = \argmax_{\messg' \in \Messgs}
\Sstrat(\state,\messg')$. If that $\Spure$ is convex, we also count
$\Sstrat$ as convex. The sender strategy in
Figure~\ref{fig:example_stratsA} is not convex, while the one in
Figure~\ref{fig:example_stratsB} is.  One would a priori not expect a
systematic relation between imprecision and convexity.

@@

\subsubsection{Results}

@@

Figure~\ref{fig:MeanMetrics} shows plots summarizing some results.  For perspicuity, we only
plot results for one level of tolerance $\beta = 0.1$, and leave out the case of $n_s =
90$. Still, every qualitative trend mentioned in the following applies to the whole set of
results.

As expected, increasing imprecision leads to higher entropy and lower
expected utility. Importantly, however, imprecision does not
necessarily lead to disastrous decline of communicative success. What
was not expected, on the other hand, was a relation between
imprecision and convexity. However, the corresponding plot clearly
shows that higher imprecision led to a higher number of outcomes with
convex sender strategies. Moreover, it also led to higher likelihood
of convergence. In fact, sufficient imprecision always ensured
convergence and convexity. It appears that perceptual imprecision
leads to more vagueness, slightly less communicative efficiency, but
more regular, well-behaved languages in shorter time.

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by imprecision. To investigate this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the distance between strategies
$\Sstrat$ and $\Sstrat'$ to be the maximum of $\text{HD}(\Sstrat,\Sstrat')$ and
$\text{HD}(\Sstrat^*,\Sstrat')$ where $\Sstrat^*$ is $\Sstrat$ with reversed message
indices. An example of the \emph{within group distance}, i.e., the average distances between
all sender strategies obtained for the same parameter values, is plotted in
Figure~\ref{fig:GroupMeasuresA} for $\beta = 0.1$ and $n_s = 10$. Despite some quantitative
differences, the general trend is the same for all other parameter settings that we tested:
with increasing imprecision, the sender strategies evolving under the \rdd were much more alike
(modulo swapping of messages). This means that perceptual imprecision unifies evolutionary
outcomes and guarantees sender strategies that are not only convex, but also regular in that
they induce a vague category split exactly in the middle of the unit interval.

@@

This unifying property of perceptual imprecision could be considered
an evolutionary beneficial side-effect. Higher imprecision can lead to
higher \emph{within group expected utility}, defined as the average
expected utility that each evolved language scored when playing against an
arbitrary other language obtained for the same parameter values. Figure~\ref{fig:GroupMeasuresB} gives a representative
example. The observation repeats for other parameter values: while
imprecision might decrease the communicative efficiency of individual
languages, it increases the conceptual coherence and communicative
success between independently evolving strategies. It is almost as if
mere confusability of states imposes a regularity constraint on
evolving categories.

@@

\section{Discussion}
\label{sec:discussion}

\subsection{Interpretation of model and results}
\label{sec:model-interpretation}

The \rdd adds expected confusion of similar states to the \rd in behavioral strategies. The
result is an abstract evolutionary dynamic that describes the most likely path of development
of aggregate population behavior that is consistent with the idea that agents imitate other
agents' behavior (Appendix~\ref{sec:deriv-rdd-imit}). This would be a process of cultural
evolution operating on behavioral strategies. But we speculate that the \rdd is also extensible
to biological evolution on mixed strategies, in which case it is a special case of the \rmd
with mutation matrices derived from confusion matrices, as shown in
Appendix~\ref{sec:diffusion-as-special}.

The \rdd is not only compatible with several interpretations of how,
at an agent level, the population-level changes in strategy
proportions come about. It is also compatible with several
interpretations of diffusion. Diffusion can be pictured as a flaw or a
virtue. If we think of perceptual imprecision as a hard limitation on
information processing, vagueness is a natural byproduct of language
evolution. Results from numerical simulations show that such
imprecision leads to vagueness, but does not need to undermine the
prospect of evolving regular and by-and-large efficient signaling
behavior. On the other hand, there appears to be a possibly beneficial
side-effect of state confusability: it speeds up convergence to
regularly shaped and convex categories. This suggests that diffusion
of behavior, as we implemented here, could also be seen as a form
of stimulus-generalization that is evolutionarily beneficial, because
it accelerates the emergence of uniform category formation. This
latter view is also suggested by a like-minded account to which we
turn next.

\subsection{Relation with previous accounts}
\label{sec:relat-with-prev}

\paragraph{Generalized reinforcement.}

% Under \emph{Herrnstein reinforcement learning}, sender and receiver
% adjust their dispositions to act after each round of play, in such a
% way that the actual (non-negative) payoff gained in the previous
% interaction is added to the non-normalized propensities for repeating
% the same behavior. For sim-max games, this means that, when the sender
% chose $\messg$ in state $\state$, and this resulted in some
% non-negative payoff, the probability that the sender chooses $\messg$
% again in $\state$ is increased, but nothing else changes. In
% particular, the sender's behavior in other choice points does not
% change. \emph{Generalized reinforcement learning} is different. When
% the use of $\messg$ in $\state$ gave positive payoff, then not only
% will the probability of its future use be promoted at $\state$ but also at
% other states, proportional to how similar these are to
% $\state$. Similar amendments take care of the way that the receiver
% updates his choice dispositions.

\citet{OConnor2013:The-Evolution-o} introduces a variant of classical Herrnstein reinforcement
learning for sim-max games, in which successful play leads to reinforcement of acts also for
states similar to the one that actually occurred. This is a form of generalization to unseen
contingencies. It not only leads to vague signaling of the appropriate kind, but also speeds up
learning in such a way that, especially for sim-max games with higher numbers of states, higher
levels of communicative success are reached in shorter learning
periods. % Technically, this result is
% partly due to the fact that signalers make bigger changes to their behavioral strategies
% after each round of play under generalized reinforcement than under the Herrnstein
% variety. That only explains the speed of adaptation, not necessarily that generalization also
% leads to regularity and communicative efficiency, but it does.

Diffusion of strategies in the \rdd can also be conceived of as a form of generalization, and
works in large part quite analogously to stimulus generalization in
\citeauthor{OConnor2013:The-Evolution-o}'s approach. But there are still
differences. \citeauthor{OConnor2013:The-Evolution-o} showed that her generalized reinforcement
learning can speed up the development of efficient signaling, especially for higher numbers of
states. Complementing this, we showed that diffusion regularizes evolutionary outcomes and
prevents meandering around sub-optimal and non-convex strategies, also for low numbers of
states.

Conceptually, the \rdd is a more abstract framework than generalized reinforcement
learning. The latter is most naturally seen as an agent-based learning dynamic that has two
players adapt their individual strategies after each concrete round of play. In contrast, the
\rdd describes a more abstract, average dynamical change in behavioral dispositions at the
aggregate population level. Although the behavior of (some forms of) reinforcement learning
mirror those of the \rd at some stage in time
(\cite{BorgersSarin997:Learning-Throug,HopkinsPosch2005:Attainability-o,Beggs2005:On-the-Converge}),
this does not mean that \emph{generalized} reinforcement learning is also necessarily a
plausible population-level description of, say, generalized learning in a population of
agents. Seen in this light, generalized reinforcement learning and the \rdd nicely complement
each other, as similarly-minded accounts operating at different levels of abstraction.
Although it would be interesting to explore the possibility of a mathematical connection
between the two, this is beyond the scope of this paper.

\paragraph{Quantal response equilibria.}
\citet{FrankeJager2010:Vagueness-Signa} suggested a number of ways in which
information-processing limitations could lead to vague strategies. The model that is most
clearly related to the present approach uses the notion of a \emph{quantal response}, also
known as a \emph{soft-max function}
\citep[e.g.][]{Luce1959:Individual-Choi,McFadden1976:Quantal-Choice-,GoereeHolt2008:Quantal-Respons}. A
quantal response is a stochastic generalization of the best response function (i.e., the
deterministic choice of an action that maximizes expected utility). Quantal responses factor in
stochastic trembles at the level of the computation or maximization of expected utilities:
agents may make mistakes in calculating how good given choice options really are for
them. \citet{FrankeJager2010:Vagueness-Signa} demonstrate that quantal response equilibria of
sim-max games can show the desired marks of vague
signaling. Figure~\ref{fig:exampleQRE_stratsA} gives an example of a quantal response
equilibrium for a sim-max game, as used in our set-up but with higher tolerance $\toler =
0.5$.
Sender and receiver strategies look very much like what evolves under \rdd with modest values
of perceptual imprecision.

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/exampleStratQRE_tolerance05.pdf}
    \caption{$\ns = 90$, $\lambda = 15$, $\toler = 0.5$}
    \label{fig:exampleQRE_stratsA}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/exampleStratQRE_tolerance005.pdf}
    \caption{$\ns = 90$, $\lambda = 15$, $\toler = 0.05$}
    \label{fig:exampleQRE_stratsB}
  \end{subfigure}

  \caption{Examples of vague quantal response equilibria. Here, $\lambda$ is an imprecision
    parameter measuring the extent of noise in the calculation of expected utilities
    \citep[see][for details]{FrankeJager2010:Vagueness-Signa}.}
  \label{fig:exampleQREs}
\end{figure}

But not all quantal response equilibria are equally plausible.  To see what the problem is,
consider Figure~\ref{fig:exampleQRE_stratsB}, which is the quantal response equilibrium for a
game with lower tolerance $\toler = 0.05$. Unlike what evolves under \rdd in this case, sender
strategies have vague boundaries also towards the end of the unit intervals. Technically, this
is because quantal responses equalize message use far away from the ``prototypical''
interpretation, not just in-between categories, so to speak. Conceptually, this is because
quantal responses introduce noise at the level of computing or maximizing expected utility of
choices.

In contrast, the \rdd implements noise at the level of the perception of different
states. Unintuitive outcomes like in Figure~\ref{fig:exampleQRE_stratsB} are thereby
avoided. Still, we believe that it would be most plausible to assume that both perceptual
confusion \emph{and} computational errors in maximizing expected utility play a role in making
vague categorization a pervasive feature of natural language meaning. Perhaps even further
factors, such as limited observations, imperfect memory or uncertainty about the exact prior
distribution of states, should be considered in order to form a more complete understanding of
the phenomenon of vagueness.

\section{Conclusions}
\label{sec:conclusions}

Vagueness is a pervasive but seemingly non-disruptive feature of
natural communication and classification systems. From an evolutionary
point of view, the challenge arises to explain how vagueness can
persist under selective pressure for precision. This is foremost a
technical challenge, probing for the possibility of integrating into a
consistent model forces that lead to vagueness and forces that lead
to efficient information transfer. This paper proposed one such model
in the context of sim-max games and explored some of its consequences.

We introduced the replicator diffusion dynamic and identified it as a special case of the
replicator mutator dynamic (Appendix~\ref{sec:diffusion-as-special}). The \rdd incorporates a
special kind of stochastic mutation due to differential confusability of similar states. It can
be thought of as the aggregate population-level behavior process that results from imitation
among agents that are unaware of perceptual confusion
(Appendix~\ref{sec:deriv-rdd-imit}). Based on data from numerical simulations, we demonstrated
that the inclusion of mild levels of perceptual imprecision does not undermine the possibility
of evolving relatively successful signaling strategies. On the contrary, strategies that
evolved under mild imprecision induce a highly regular and systematic category structure that
shows the signs of vagueness as desired. There might even be a higher-order benefit to the
presence of imprecision, in that it can accelerate convergence to optimal categorization, by
preventing evolutionary paths to stay near inefficient non-convex strategies for a long time. A
little imprecision thus also unifies and regularizes evolutionary outcomes in such a way that
sub-optimal categorization is avoided in short-term trajectories.

>>>>

\begin{abstract}
  Signaling games are popular models for studying the evolution of meaning, but typical
  approaches do not incorporate vagueness as a feature of successful signaling.  Complementing
  recent like-minded models, we describe an aggregate population-level dynamic that describes a
  process of imitation of successful signaling behavior under imprecise perception and
  realization of similar stimuli. Applying this new dynamic to a generalization of Lewis'
  signaling games, we show that stochastic imprecision leads to vague, yet by-and-large
  efficient signal use, and, moreover, that it unifies evolutionary outcomes and helps avoid
  suboptimal categorization. The upshot of this is that we see \emph{as-if}-generalization at
  an aggregate level, without agents actually generalizing.
\end{abstract}

@@

A number of authors have recently tried to explain why vagueness evolved as something that is
itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing \citep[e.g.][]{FrankeJager2010:Vagueness-Signa} or the result of particular
low-level learning strategies \citep[e.g.][]{OConnor2013:The-Evolution-o}. This paper
contributes to the latter line of thought. Concretely, we introduce a dynamic that describes
average behavior resulting from imitation under confusability of similar stimuli. We show that
stochastic noise can not only lead to vague, yet relatively efficient signal use, but can also
unify evolutionary outcomes and help avoid suboptimal categorization: regularities in
confusability of stimuli also regularize signaling behavior and evolving categories. The system
gives rise to generalization behavior without agents actually generalizing.

@@

How does a tendency to confuse similar states interact with selective pressure towards more
efficient signaling strategies under the imprecise conditional imitation dynamic? If confusion
probabilities are moderate and regular, in that they track similarity of states, we might
expect a regularizing effect also on evolving signaling strategies. Indeed, we hypothesize that
confusion of states can give rise to population-level aggregate behavior that looks \emph{as
  if} information about what is good for one state percolates also to similar states. In other
words, imprecise imitation may make signaling behavior on average look like agents generalize
across similar states, even if no agent actually generalizes. To explore whether confusion of
states can have this effect, we turn to numerical simulation.

\subsection{Setting the stage}
\label{sec:setting-stage}

To obtain concrete results, we must fix how to represent states, similarity of states, the
conditional confusion probabilities $P_o$ and $P_r$ and the utilities of our sim-max
games. Confusion probabilities between states should be a function of perceptual similarity:
the more similar two states are, the more likely they could be mistaken for each other.

Let the state space consist of $n \ge 2$ states that are equally spaced across the unit
interval, including $0$ and $1$. All states occur, for simplicity, with the same probability.
The distance $\card{\mystate{i} - \mystate{j}}$ is the objective, physical similarity between
two states $\mystate{i}$ and $\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by \citet{Nosofsky1986:Attention-Simil}:
\begin{align*}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \imprecision) =
      \begin{cases}
    1 & \textrm{if } \imprecision = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \imprecision = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \imprecision^2} \right ) & \textrm{otherwise,} \\
    \end{cases}
\end{align*}
where $\imprecision \ge 0$ is an imprecision parameter. When $\imprecision=0$ agents perfectly
discriminate between states; when $\imprecision \rightarrow \infty$ agents cannot discriminate
states at all. Figure~\ref{fig:NosofskySim} gives an impression of Nosofsky-similarity for
different parameter values. Other formalizations of perceptual similarity are possible,
including ones that allow for different discriminability in different areas of the state space,
but we stick with Nosofsky's similarity function for the time being, because it is
mathematically simple, yet an established notion in mathematical psychology.

@@

We further assume that the probability of confusing any two states $\mystate{i}$ and
$\mystate{j}$ is proportional to their perceived similarity and, to keep matters simple, that
observational confusability $P_o$ is simply the same as realizational confusability $P_r$ and
that both are governed by the same imprecision parameter $\alpha$:
\begin{align*}
  P_o(\mystate{o} \probbar \mystate{a}) & \propto \similarity(\mystate{o}, \mystate{a} \myts ; \myts
  \imprecision) &   P_r(\mystate{r} \probbar \mystate{i}) & \propto \similarity(\mystate{r}, \mystate{i} \myts ; \myts
  \imprecision) \,.
\end{align*}
For $\imprecision = 0$, we obtain trivial confusion probabilities for which everything is reduced
to the replicator dynamic. For $\imprecision > 0$, any state can be confused for any other
state with some positive probability. For the uninteresting case of
$\imprecision \rightarrow \infty$, confusion is maximal and every state can be perceived or
realized as any other state with probability $\nicefrac{1}{\card{\States}}$.

@@

In sum, to structure thinking about the behavior of our imprecise imitation dynamic, the system
is governed by three parameters: the number $\ns = \card{T}s$ of states in the state space,
imprecision $\imprecision$ and tolerance $\toler$.

\subsection{Experimental set-up}
\label{sec:simulations}

We ran 50 trials of the discrete-time dynamic in (\ref{eq:nRD}), starting with randomly sampled
sender and receiver strategies, for each triplet of independent parameter values:
$\ns \in \set{6, 10, 50, 90}$, $\imprecision \in \set{0, 0.05, 0.1, 0.2, 0.3}$,
$\toler \in \set{0.05, 0.1, 0.2, 0.3}$. Each trial ran for a maximum of 200 update steps of the
\rdd. A trial was considered converged, and thus stopped before the maximum of 200 rounds, if
the total amount of change between strategies before and after an update step was smaller than
a suitably chosen threshold. It is not guaranteed that strategies at halting time had converged
to the eventual attracting state, whether they ran for 200 rounds or not. Our notion of
convergence is therefore only a categorical measure for reaching a certain (well-considered,
but eventually arbitrary) degree of stability. In other words, out notion of ``convergence'' is
a measure of relative speed: it is true that the system reached a state in which evolutionary
adaptations had slowed down almost to a halt before 200 update steps? This is motivated by
pragmatic concerns regarding length of simulation time, but also theoretically justified,
because we hypothesize that confusability of states leads to regularization of evolving
strategies, which would show exactly in an increased speed of evolutionary trajectories towards
well-behaved and regular signaling behavior.

@@

Representative examples for resulting strategy pairs are given in
Figure~\ref{fig:example_strats}. Figure~\ref{fig:example_stratsA} shows a strategy pair at
stopping time with 90 states, tolerance $\toler = 0.1$ and imprecision $\imprecision = 0$. Zero
imprecision means that the trial was effectively an application of the plain \rd. Noteworthily,
the given sender strategy approximates a pure sender strategy that crisply partitions the state
space into non-convex sets. The irregular shape of the receiver strategy suggests that the
pictured strategy pair has not yet reached a dynamically stable state under the \rd. Indeed,
the trial was stopped when reaching the maximum of 200 rounds. In contrast, the outcome of a
trial with identical parameters, except with imprecision $\imprecision = 0.05$, which is shown
in Figure~\ref{fig:example_stratsB}, had converged (in our technical sense) after 99
rounds. The sender strategy shows a smooth blending from one ``category'' to the other, and the
receiver's interpretations are rather extended curves, peaking at a central point in the
relevant ``categories.''

@@

\subsection{Dependent measures}

@@

\paragraph{Convexity.} At least for sender strategies, which develop faster than receiver
strategies, it also makes sense to define a categorical measure of convexity that compensates
for potential vagueness. To determine whether a sender strategy $\Sstrat$ is convex despite
possibly being vague, we look at the derived pure strategy $\Spure$ for which
$\Spure(\state) = \argmax_{\messg' \in \Messgs} \Sstrat(\state,\messg')$. If that $\Spure$ is
convex, we also count $\Sstrat$ as convex. The sender strategy in
Figure~\ref{fig:example_stratsA} is not convex, while the one in
Figure~\ref{fig:example_stratsB} is.  If confusion of states can regularize signaling
strategies, like we hypothesized, we should see more convexity with increasing impairment all
else equal.

@@

\subsection{Results}

@@

Figure~\ref{fig:MeanMetrics} shows plots summarizing a selected part of our findings.  For
perspicuity, we only plot results for one level of tolerance $\beta = 0.1$, and leave out the
case of $n_s = 90$. Still, every qualitative trend mentioned in the following applies to the
whole set of results.

As expected, increasing imprecision leads to higher entropy and lower expected
utility. Importantly, however, imprecision does not necessarily lead to disastrous decline of
communicative success. What is more, in line with our hypothesis that mere imprecision in
imitation behavior can lead to behavior that looks like agents generalize across similar
stimuli, higher imprecision led to a higher number of outcomes with convex sender
strategies. It also led to higher rates of convergence. In fact, sufficient imprecision always
ensured convergence and convexity. It appears that perceptual imprecision leads to more
vagueness, slightly less communicative efficiency, but more regular, well-behaved languages in
shorter time.

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by imprecision. To investigate this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the distance between strategies
$\Sstrat$ and $\Sstrat'$ to be the maximum of $\text{HD}(\Sstrat,\Sstrat')$ and
$\text{HD}(\Sstrat^*,\Sstrat')$ where $\Sstrat^*$ is $\Sstrat$ with reversed message
indices. An example of the \emph{within group distance}, i.e., the average distances between
all sender strategies obtained for the same parameter values, is plotted in
Figure~\ref{fig:GroupMeasuresA} for $\beta = 0.1$ and $n_s = 10$. Despite some quantitative
differences, the general trend is the same for all other parameter settings that we tested:
with increasing imprecision, the resulting sender strategies were much more alike (modulo
swapping of messages). This means that perceptual imprecision can speed up and unify
evolutionary outcomes. It can amplify the emergence of sender strategies that are not only
convex, but also regular in that they induce a vague category split exactly in the middle of
the unit interval.

@@

This unifying property of perceptual imprecision could be considered an evolutionary beneficial
side-effect. Higher imprecision can lead to higher \emph{within group expected utility},
defined as the average expected utility that each evolved language scored when playing against
an arbitrary other language obtained for the same parameter
values. Figure~\ref{fig:GroupMeasuresB} gives a representative example. The observation repeats
for other parameter values: while imprecision might decrease the communicative efficiency of
individual languages, it increases the conceptual coherence and communicative success between
independently evolving strategies. It is as if mere confusability of states imposes a
regularity constraint on evolving categories.

@@

\section{Conclusion}
\label{sec:discussion}

We set out to meet a technical challenge posed by Lipman's problem
\citep{Lipman2009:Why-is-Language}: is there a conceptually sound and mathematically coherent
formal model that shows how vague language can evolve under selective pressure for efficient
communication? On the assumption that agents are likely to mistake states for one another,
especially when they are perceptually similar, derived a generalization of the replicator
dynamic in terms of imitation of successes. The resulting population-level dynamic produced
signaling behavior that is at the same time regular and by-and-large communicatively efficient,
while also showing the crucial marks of vagueness. In a sense, the model derives vagueness as a
by-product of an arguably natural limitation on the discriminatory power of signaling agents.

But the inclusion of confusability of similar states also had noteworthy effects on the
evolving meaning of signals. It transpired that imprecision can have further accelerating and,
surprisingly, unifying effects on meaning evolution. State confusability has an effect on
aggregate signaling behavior that can be described as generalization without generalizers: the
dynamics of imprecise imitation look as if ``conclusions'' about what works for one state are
``carried over'' to similar states. This, however, is merely an epiphenomenon in the sense no
single agent genuinely generalizes over stimuli or reasons about what a more systematic
signaling strategy would be.

In sum, inability to sharply discriminate similar stimuli may lead to vagueness and bring about
a (slight) decrease in communicative efficiency. But there may also be an advantage, not of
vagueness itself, but of its reason: systematicity in the confusability of states (which may be
a natural by-product of the perceptual system) supports \emph{as-if}-generalization at the
population-level without having to assume that agents themselves have any generalization power.

\citet{OConnor2013:The-Evolution-o} makes a related case based on a version of reinforcement
learning for sim-max games, in which successful play leads to reinforcement of choice options
also for states similar to the ones that actually occurred. This not only leads to vague
signaling of the appropriate kind, but also speeds up learning in such a way that, especially
for sim-max games with higher numbers of states, higher levels of communicative success are
reached in shorter learning periods. Our results complement and extend
\citeauthor{OConnor2013:The-Evolution-o}'s. The most important differences are that we obtain
similar regularizing effects also for cases with low numbers of states and do not assume that
agents have any kind of generalizing capacity in and of themselves, even if that is only
implicit, as in \citeauthor{OConnor2013:The-Evolution-o}'s generalized reinforcement learning.

\citet{FrankeJager2010:Vagueness-Signa} suggested a number of ways in which
information-processing limitations could lead to vague strategies. The model that is most
clearly related to the present approach uses the notion of a \emph{quantal response}, also
known as a \emph{soft-max function}
\citep[e.g.][]{Luce1959:Individual-Choi,McFadden1976:Quantal-Choice-,GoereeHolt2008:Quantal-Respons}. The
main difference between this and our present approach is in where stochastic noise is assumed
to reside. In case of a quantal response dynamic, it is the computation of expected utilities;
in case of imprecise imitation, it is perception and realization of similar states. There are
cases, then, where evolving signaling behavior, as predicted by these two approaches, is quite
different. Intuitively speaking, for a case with two messages, the further we venture away from
a prototypical interpretation of either message, the less discriminative a signaling strategy
would be when the source of ``trembles'' is the computation of expected utilities: to wit,
since both ``tall'' and ``short'' are almost equally bad descriptions for a giant, quantal
response dynamics predict that speaker's would be almost indifferent. Speaker behavior that
evolves under confusion of states does not have this puzzling property.
